---
title: "Attenuated, exaggerated, or flipped health disparities: Ramifications of measurement error on concentration indices"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 6, fig.height = 4)
```

This repository contains `R` code to reproduce results from the manuscript by Hung, Green, and Lotspeich (2025+). 

## Packages 

Our implementation of the health disparities methods in the paper rely on the following `R` packages, which can be installed and loaded as follows. 

```{r}
# Run once: install.packages(c("dplyr", "ggplot2", "latex2exp", "scales"))
library(dplyr) ## for data wrangling
library(ggplot2) ## for plots 
library(latex2exp) ## for LaTeX in plots
library(scales) ## for percent labels in plots
library(ggconc) ## for concentration curves

# Run once: devtools::install_github("sarahlotspeich/mochi")
library(mochi) ## for moment-based correction
```

## Electronic Health Records Data: Diabetes and Proximity to Healthy Foods 

We begin with the electronic health records (EHR) data from our previous work, which already has diabetes diagnosis (yes/no) and proximity to healthy foods (both error-free and error-prone). With permission from the Institutional Review Board at Wake Forest University School of Medicine, this de-identified version of the study dataset was made available with our prior work as Supplementary Material. These data can be loaded as follows. 

```{r}
# Read in the data from CSV
ehr_food = read.csv("https://raw.githubusercontent.com/sarahlotspeich/ME_Disparities_HSORM/refs/heads/main/me_disparities_analysis_2020.csv") |> 
  select(ID, MAP_PROXIMITY, STRAIGHT_PROXIMITY)

## Inspect data summary
ehr_food |> 
  summary()
```

We retain the following columns from the full electronic health records (EHR) dataset. 

  - `ID`: a deidentified patient identifier 
  - `MAP_PROXIMITY`: error-free, map-based proximity to healthy foods (in miles)
  - `STRAIGHT_PROXIMITY`: error-prone, straight-line proximity to healthy foods (in miles)
  
Each row represents data for one patient. We also add columns with the fractional ranks for all patients based on `MAP_PROXIMITY` and `STRAIGHT_PROXIMITY`. 

```{r}
# Add columns for fractional ranks based on error-prone and error-free X
ehr_food = ehr_food |> 
  mutate(STRAIGHT_RANK = (rank(-STRAIGHT_PROXIMITY) - 1) / n() + 1 / (2 * n()), 
         MAP_RANK = (rank(-MAP_PROXIMITY) - 1) / n() + 1 / (2 * n()))
```

## Monte Carlo Simulation: Generating Hemoglobin A1c

For the current work, we wanted to simulate hemoglobin A1c (HbA1c, a biomarker for blood glucose that is key for monitoring type-2 diabetes) for these patients. The goal was to demonstrate how these patients' error-prone versus error-free proximity to healthy foods (from the real EHR data) may tell a different story about potential disparities in HbA1c. In the code chunk below, we demonstrate how HbA1c was generated for a single Monte Carlo simulation, where the targeted magnitude of the disparity was $C \approx 0.5$.  

```{r}
# Set parameters to generate Y|R
beta0 = 2.5     ## average HbA1c for patients with optimal proximity (0 miles)
beta1 = -3.0    ## coefficient on fractional rank based of proximity (in miles)

# Add column with generated Y|R
set.seed(918) ## for reproducibility
ehr_food = ehr_food |> 
  mutate(HBA1C = rnorm(n = n(), 
                       mean = beta0 + beta1 * MAP_RANK, 
                       sd = 1))

## Inspect data summary
ehr_food |> 
  summary()
```

## Exploratory Data Analysis 

```{r}
ehr_food |> 
  ggplot(aes(x = MAP_PROXIMITY, y = STRAIGHT_PROXIMITY)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  coord_equal() + 
  theme_minimal(base_size = 14) + 
  labs(x = TeX("Map-Based Proximity $X$"), 
       y = TeX("Straight-Line Proximity $X^*$"))
```

```{r}
ehr_food |> 
  ggplot(aes(x = MAP_RANK, y = STRAIGHT_RANK)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  coord_equal() + 
  theme_minimal(base_size = 14) + 
  labs(x = TeX("Fractional Rank $R$ Based on Map-Based Proximity $X$"), 
       y = TeX("Fractional Rank $R^*$ Based on Straight-Line Proximity $X^*$"))
```

## Analysis 

### Concentration Curves

```{r}
ehr_food |> 
  ggplot() +
  geom_concentration(aes(x = MAP_PROXIMITY, y = HBA1C, color = "Driving Distance\n(Error-Free)"), 
                      rank_ascend = FALSE, round_rank = 0) +
  geom_concentration(aes(x = STRAIGHT_PROXIMITY, y = HBA1C, color = "Haversine Distance\n(Error-Prone)"), 
                      rank_ascend = FALSE, round_rank = 0) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  scale_color_manual(values = c("Driving Distance\n(Error-Free)" = "#39558CFF", 
                                "Haversine Distance\n(Error-Prone)" = "#E85362FF"), 
                     name = "Proximity Based on") +
  coord_equal() +
  theme_minimal() +
  labs(x = "Cumulative Proportion of Patients\n(Ranked by Proximity)",
       y = "Cumulative Proportion of Hemoglobin A1c") + 
  theme(axis.title = element_text(face = "bold"), 
        legend.title = element_text(face = "bold")) + 
  scale_x_continuous(labels = label_percent()) + 
  scale_y_continuous(labels = label_percent())
```

### Concentration Index

```{r}
## Concentration index based on straight-line proximity (unbinned)
ci_straight = ci(ineqvar = ehr_food$NEG_STRAIGHT_PROXIMITY, ### Food environment exposure
                 outcome = ehr_food$HBA1C, ### Health outcome 
                 type = "CIw", ### Wagstaff's correction for binary health outcome
                 robust_se = TRUE) ### Use sandwich standard errors for confidence interval 
summary(ci_straight)

## Concentration index based on map-based proximity (unbinned)
ci_map = ci(ineqvar = ehr_food$NEG_MAP_PROXIMITY, ### Food environment exposure
                 outcome = ehr_food$HBA1C, ### Health outcome 
                 type = "CIw", ### Wagstaff's correction for binary health outcome
                 robust_se = TRUE) ### Use sandwich standard errors for confidence interval 
summary(ci_map)
```